---
title: "Predator_Conservation_EDA"
author: "Khadijah Amarah Moosa"
date: "2025-02-20"
output: html_document
---


## Configuring knitr
I set up a R markdown document by configuring knitr options.
I did this to ensure that my R code is visible in my final document but my setup chunk is hidden. 

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load essential R libraries
I loaded essential R libraries for data handling and visualization.
I did this to ensure my necessary libraries are available for data processing and visualizing in my R session.

```{r loading libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
```


## Load raw data into R for analysis
I read the csv file into R using the read.csv function.
I did this to ensure that the raw data is loaded into R for analysis.

```{r reading file}
raw_data <- read.csv("C:/Users/Student/Documents/GitHub Deliverable/predator_conservation/NonJournalRawDataIN.csv", sep = ",")

```


## Perform inital checks on imported dataset
I performed initial checks on the imported dataset. 
I did this to confirm that the dataset is corretly imported and to get an overview of its structure before further processing.

```{r EDA}

# check top 5 rows to ensure data loaded correctly 
head(raw_data, 5)

# check column names to see if tidy
colnames(raw_data)

```


## Tidy dataset
I tidied the dataset by renaming columns to more readable and consistent names.
I did this to improve the readability of the data and to make subsequent analysis easier to work with.

```{r tidy data}

# rename columns
tidy_data <- raw_data %>%
  rename(
    Paper = `Paper`,
    Journal = `Journal..1..non.journal.0.`,  # Renaming for clarity
    Study_Design = `Study.design`,
    Predator_Species = `Predator.species`,
    Protected_Assets = `Protected.assets`,
    Intervention_Category = `Category.of.intervention`,
    Intervention_Type = `Type.of.intervention`,
    Intervention_Details = `Intervention.details`,
    Country = `Country`,
    Damage_Metric = `Damage.metric`,
    Relative_Risk_RR = `Relative.risk.RR`
  )

# view the renamed dataset
head(tidy_data)

```


## Clean data and prepare for further analysis
I separated multiple values into different rows and I converted various columns
to appropriate data types. I also checked to structure of the dataset after the transformations and 
confirmed the changes that I made to the column types.
I did these steps to clean the data and prepare it for further analysis.

```{r}

# Separate multiple values into different rows
tidy_data <- tidy_data %>%
  separate_rows(Predator_Species, sep = ", ") %>%
  separate_rows(Protected_Assets, sep = ", ")

# View the transformed dataset
head(tidy_data)

# Convert column types
tidy_data <- tidy_data %>%
  mutate(
    Paper = as.character(Paper),  # Convert Paper names to character
    Journal = as.factor(Journal),  # Convert Journal column to factor (categorical)
    Study_Design = as.factor(Study_Design),  # Convert study design to factor
    Predator_Species = as.factor(Predator_Species),  # Convert predator species to factor
    Protected_Assets = as.factor(Protected_Assets),  # Convert protected assets to factor
    Intervention_Category = as.factor(Intervention_Category),  # Convert intervention category to factor
    Intervention_Type = as.factor(Intervention_Type),  # Convert intervention type to factor
    Intervention_Details = as.character(Intervention_Details),  # Keep as character
    Country = as.factor(Country),  # Convert country to factor
    Damage_Metric = as.character(Damage_Metric),  # Keep as character (might need more cleaning)
    Relative_Risk_RR = as.numeric(Relative_Risk_RR)  # Convert to numeric for calculations
  )

# Check structure of the updated dataset
str(tidy_data)
```


## Create summary dataset and visualize relationship of the dataset
I created a summary dataset and visualized the relationship between Study Design and Relative Risk (RR), grouped by journal type. I created a scatter plot with  error bars.

```{r}
summary_data <- tidy_data %>%
  group_by(Study_Design, Journal) %>%
  summarise(
    median_RR = median(Relative_Risk_RR),
    min_RR    = min(Relative_Risk_RR),
    max_RR    = max(Relative_Risk_RR),
    count     = n(),
    .groups   = "drop"
  )

ggplot(summary_data, aes(x = Study_Design, y = median_RR, color 
                         = factor(Journal))) +
  geom_point(
    position = position_dodge(width = 0.5),
    size = 4
  ) +
  geom_errorbar(
    aes(ymin = min_RR, ymax = max_RR),
    position = position_dodge(width = 0.5),
    width = 0.2,
    size = 1
  ) +
  # ─────────────────────────────────────────────────────────────
  # Add text for count at the top of each error bar
  geom_text(
    aes(label = count, y = max_RR),             # Place label at max_RR
    position = position_dodge(width = 0.5),     # Same dodge
    vjust = -0.5,                               # Slightly above the error bar
    size = 3,                                   # Text size
    color = "black"                             # Text color (can also map to Journal)
  ) +
  # ─────────────────────────────────────────────────────────────
  scale_color_manual(
    values = c("red", "blue"),
    labels = c("Non-journals", "Journals")
  ) +
  labs(
    title = "Study Design vs. Relative Risk (RR)",
    x = "Study Design",
    y = "Relative Risk (RR)",
    color = "Journal Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```


Figure 1: This graph illustrates how the median Relative Risk (RR) varies by  Study Design and it also determines whether the study is from a journal or  non-journal source. The error bars represent the range of minimum to maximum RR, and the counts of studies are labelled at the top of the bars.